My Title

### 2023-10-14 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw
Adding background pic.
Update cheatsheet page 1 for v2.0.

@@

@@

{
     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
     \vspace{2mm}
}

\begin{multicols*}{4}
%\setlength{\premulticols}{1pt}
%\setlength{\postmulticols}{1pt}
%\setlength{\multicolsep}{1pt}
%\setlength{\columnsep}{2pt}

@@

It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts}, \emph{sf} and \emph{plm} classes for panel data ('pseries', 'pdata.frame').  \\ [0.8em]

\emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities. Its default is \code{na.rm = TRUE}, and implemented at very low cost at the algorithm level.  \\ [0.8em]

@@

\quad \code{fmean, fmedian, fmode, fsum, fprod, fsd, fvar} \\
\quad \code{fmin, fmax, fnth, ffirst, flast, fnobs, fndistinct} \newline

@@

\vspace{2mm}
\textbf{Examples}
<<>>=
fmean(AirPassengers)  # Vector
fmean(AirPassengers, w = cycle(AirPassengers))  # Weighted mean
fmean(EuStockMarkets) # Matrix
fmean(EuStockMarkets, drop = FALSE) # Don't drop dimensions
fmean(airquality)     # Data Frame (can also use drop = FALSE)
fmean(iris[1:4], g = iris$Species) # Grouped
X = iris[1:4]; g = iris$Species; w <- abs(rnorm(nrow(X)))
fmean(X, g, w)  # Grouped and weighted (random weights)
## Transfomrations: here centering data on the weighted group median
TRA(X, fmedian(X, g, w), "-", g) |> head(3)
fmedian(X, g, w, TRA = "-") |> head(3) # Same thing: more compact
fmedian(X, g, w, "-", set = TRUE) # Modify in-place (same as setTRA())
head(iris, 3) # Changed iris too, as X = iris[1:4] did a shallow copy
@

% \begin{addmargin}[2em]{0em}
% \code{fmean(data[3:5], data\$grp1, data\$weights)\\
% data \%>\% fgroup\_by(grp1) \%>\% fmean(weights)\\
% TRA(mat, fmedian(mat, g), "-", g)\\
% fmedian(mat, g, TRA = "-")  \# same thing
% }
% \end{addmargin}

@@

\section{Grouping and Ordering}
% \vspace{1mm}
\itxt{Optimized functions for grouping, ordering, unique values, splitting \& recombining, and dealing with factors}
\newline

@@

<<>>=
mtcars |> fgroup_by(cyl, vs, am) |> ss(1:2)
# Group Stats: [N. groups | mean (sd) min-max of group sizes]
# Fast Functions also have a grouped_df method: here wt-weighted medians
mtcars |> fgroup_by(cyl, vs, am) |> fmedian(wt) |> head(3)
@
%\qquad {\scriptsize \textcolor{darkgray}{\emph{Group Stats:} N. groups $|$ Mean (Std. Dev.) Min-Max of group sizes}} \newline

\code{GRPN(), fgroup\_vars(), fungroup()} - get group count,\\ \qquad grouping columns/variables, and ungroup data\\ [0.5em]
\code{qF(), qG()} - quick \code{as.factor}, and vector grouping object\\ \qquad of class 'qG': a factor-light without levels attribute\\
\setstretch{1.5}
\code{group()} - (multivariate) group id ('qG') in appearance order\\
\code{groupid()} - run-length-type group id ('qG')\\
\code{seqid()} - group-id from integer-sequences ('qG')\\
% \code{timeid()} - time variable to integer by GCD (class 'qG')\\
\code{radixorder[v]()} - (multivariate) radix-based ordering\\
\code{finteraction()} - fast factor interactions (or return 'qG')\\
\code{fdroplevels()} - fast removal of unused factor levels\\
\code{f[n]unique()} - fast unique values / rows (by columns)\\
\code{gsplit()} - fast  splitting vector based on 'GRP' objects\\
\code{greorder()} - efficiently reorder \code{y = unlist(gsplit(x, g))}\\ \qquad such that \code{identical(greorder(y, g), x)}
\setstretch{1}
\newline

@@

<<cache=TRUE>>=
X <- iris[1:4]; v <- as.character(iris$Species)
f <- qF(v, na.exclude = FALSE) # Adds 'na.included' class: no NA checks
gv <- group(v) # 'qG' object: first appearance order, with 'na.included'
microbenchmark(fmode(X, v), fmode(X, f), fmode(X, gv), fmode(X, g))
@
% \vspace{-2mm}

\hrrule
\section{Quick Conversions}
% \vspace{1mm}
\itxt{Fast and exact conversion of common data objects} \\ [0.5em]
\code{qM(), qDF(), qDT(), qTBL()} - convert vectors, arrays, data.frames or lists to matrix, data.frame, data.table or tibble\\ [0.5em]
\code{m[r|c]tl()} - matrix rows/cols to list, data.frame or data.table\\ [0.5em]
\code{qF(), as\_numeric\_factor(), as\_character\_factor()} - convert to/from factors or all factors in a list / data.frame

@@

\section{Fast Data Manipulation}
\vspace{-1.4mm}
\itxt{Efficient implementations of data manipulation functions} % \newline
\setstretch{1.5}
\code{fselect[<-]()} - select/replace columns\\
\code{fsubset()} - subset data (rows and columns)\\
\code{ss()} - fast alternative to \code{[}, particularly for data frames\\
\code{[row|col]order[v]()} - reorder (sort) rows and columns\\
\code{fmutate(), fsummarise()} - \emph{dplyr}-like, incl. \code{across()} feature\\
\code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
\code{fcompute[v]()} - compute new cols dropping existing ones\\
\code{[f|set]rename()} - rename (any object with 'names' attribute)\\
\code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
\code{get\_vars[<-]()} - select/replace columns (standard eval.)\\ [0.5em]
\setstretch{1}
\code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
\code{add\_vars[<-]()} - add or column-bind columns \newline

\textbf{Examples}
<<>>=
mtcars |> fsubset(mpg > fnth(mpg, 0.95), disp:wt, cylinders = cyl)
mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
i <- base::invisible # These are equivalent, the second option is faster:
mtcars |> fgroup_by(cyl, vs, am) |> fmutate(sum_mpg = fsum(mpg)) |> i()
mtcars |> fmutate(sum_mpg = fsum(mpg, list(cyl, vs, am), TRA = 1)) |> i()
# These are also equivalent (weighted means), again the second is faster
mtcars |> fgroup_by(cyl) |> fmutate(across(disp:drat, fmean, wt)) |> i()
mtcars |> ftransformv(disp:drat, fmean, cyl, wt, 1, apply = FALSE) |> i()
# ftransform()/fcompute() support list input and ignore attached groupings
mtcars %>% fgroup_by(cyl) %>% ftransform(fselect(., hp:qsec) %>%
           fmedian(TRA = 1) %>% fungroup() %>% fsum(TRA = "/")) |> i()
# Again a faster equivalent: note the use of 'set' to avoid a deep copy
mtcars %>% ftransform(fselect(., hp:qsec) %>% fmedian(cyl, TRA = 1) %>%
                      fsum(TRA = "/", set = TRUE)) %>% i()
# Aggregation: weighted standard deviations
mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
# Grouped linear models: .apply = FALSE applies functions to DF subset
qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(across(disp:drat,
     function(x) list(models = list(lm(disp ~., x))), .apply = FALSE))
# Adding some columns. Use ftransform<- to also replace existing ones
add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("perc_")
@
<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
% \vspace{-2mm}

@@

<<>>=
# Population weighted mean (PCGDP, LIFEEX) & mode (country), and sum(POP)
collap(wlddev, country + PCGDP + LIFEEX ~ income, w = ~ POP)
@
%\vspace{-10mm}

% \newpage

% ------------------------------------------------------------------
% Second Page
% ------------------------------------------------------------------

@@

\itxt{Namespace Masking}\\ [0.5em]
Can set \code{option(collpse\_mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R or \emph{dplyr}. A few keywords exist to mask multiple functions, see \code{help("collapse-options")}. This allows clean \& fast code, but poses additional namespace challenges:

@@

\end{multicols*}
\end{document}

>>>>

% https://tex.stackexchange.com/questions/46280/how-to-create-a-background-image-on-title-page-with-latex
\usepackage{eso-pic}
\newcommand\BackgroundPic{%
\put(0,0){%
\parbox[b][\paperheight]{\paperwidth}{%
\vspace{-3mm} % \vfill
\flushright
\includegraphics[width=0.3\paperwidth,%
keepaspectratio]{"background_cropped.png"}%
\vfill
}}}

\newcommand\Logo{%
\put(0,0){%
\parbox[b][\paperheight]{\paperwidth}{%
\vspace{-1mm} % \vfill
\flushright
\includegraphics[width=0.08\paperwidth,%
keepaspectratio]{"collapse_logo_vsmall.png"} \hspace{3mm} %
\vfill
}}}

@@

% \makebox[0pt][r]{%
%   \raisebox{-\totalheight}[0pt][0pt]{%
%     \includegraphics[width=4in]{background}}}%
\AddToShipoutPicture*{\BackgroundPic}
\AddToShipoutPicture*{\Logo}

@@

{
     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ \textcolor{darkgray}{: : CHEAT SHEET}} %\\%\small{by Sebastian Krantz} %
     % \vspace{2mm}
}

%\begin{adjustbox}{totalheight=0.5\textheight} % -2\baselineskip
% \resizebox*{!}{\textheight}{%
\begin{multicols}{4}
%\setlength{\premulticols}{1pt}
%\setlength{\postmulticols}{1pt}
%\setlength{\multicolsep}{1pt}
%\setlength{\columnsep}{2pt}

@@

It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with base R, \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts/zoo}, \emph{sf}, and \emph{plm} classes for panel data.  \\ [0.8em]

\emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities. It is globally configurable using \code{set\_collapse()} which includes algorithm defaults, multithreading, and the exported namespace (see below).  \\ [0.8em]

@@

\quad \code{fmean, fmedian, fmode, fsum, fprod, fsd, fvar,} \\
\quad \code{fmin, fmax, fnth, ffirst, flast, fnobs, fndistinct} \newline

@@

\vspace{2mm}
\textbf{Examples}
<<>>=
fmean(AirPassengers)  # Vector
fmean(AirPassengers, w = cycle(AirPassengers))  # Weighted mean
fmean(EuStockMarkets) # Matrix
fmean(airquality)     # Data Frame (use drop = FALSE to keep frame)
fmean(iris[1:4], g = iris$Species) # Grouped
X = iris[1:4]; g = iris$Species; w <- abs(rnorm(nrow(X)))
fmean(X, g, w)  # Grouped and weighted (random weights)
## Transfomrations: here centering data on the weighted group median
TRA(X, fmedian(X, g, w), "-", g) |> head(2)
fmedian(X, g, w, TRA = "-") |> head(2) # Same thing: more compact
fmedian(X, g, w, "-", set = TRUE) # Modify in-place (same as setTRA())
@
% \begin{addmargin}[2em]{0em}
% \code{fmean(data[3:5], data\$grp1, data\$weights)\\
% data \%>\% fgroup\_by(grp1) \%>\% fmean(weights)\\
% TRA(mat, fmedian(mat, g), "-", g)\\
% fmedian(mat, g, TRA = "-")  \# same thing
% }
% \end{addmargin}
\vspace{-1mm}
\hrrule
\section{Other Statistical Functions}
% \vspace{1mm}
\itxt{Fast (weighted) sample quantiles, range, and distances}\\
\setstretch{1.5}
\code{fquantile(x, probs, w, o, na.rm = TRUE, type = 7)}\\
\code{frange(x, na.rm = TRUE)} \\
\code{fdist(x, v, method = "euclidean", nthreads = 1)}\\
\setstretch{1}

@@

\section{Grouping and Ordering}
% \vspace{1mm}
\itxt{Optimized functions for grouping, ordering, unique values, matching, splitting, and dealing with factors}
\newline

@@

<<>>=
mtcars |> fgroup_by(cyl, vs, am) |> ss(1:2)
# Group Stats: [N. groups | mean (sd) min-max of group sizes]
# Fast Functions also have a grouped_df method: here wt-weighted medians
mtcars |> fgroup_by(cyl, vs, am) |> fmedian(wt) |> head(2)
@
%\qquad {\scriptsize \textcolor{darkgray}{\emph{Group Stats:} N. groups $|$ Mean (Std. Dev.) Min-Max of group sizes}} \newline

\code{GRPN(), fcount[v](), fgroup\_vars(), fungroup()} - get group count, grouping columns, and ungroup data\\ [0.5em]
\code{qF(), qG()} - quick \code{as.factor}, and vector grouping object\\ \qquad of class 'qG': a factor-light without levels attribute\\
\setstretch{1.5}
\code{group()} - (multivariate) group id ('qG') in appearance order\\
\code{groupid()} - run-length-type group id ('qG')\\
\code{seqid()} - group-id from integer-sequences ('qG')\\
% \code{timeid()} - time variable to integer by GCD (class 'qG')\\
\code{radixorder[v]()} - (multivariate) radix-based ordering\\
\code{finteraction()} - fast factor interactions (or return 'qG')\\
\code{fdroplevels()} - fast removal of unused factor levels\\
\code{f[n]unique(), fduplicated()} - fast unique values / rows\\
\code{fmatch(), \%[!][i]in\%} - fast matching of values / rows\\
\code{gsplit()} - fast  splitting vector based on 'GRP' objects\\
\code{greorder()} - efficiently reorder \code{y = unlist(gsplit(x, g))}\\ \qquad such that \code{identical(greorder(y, g), x)}
\setstretch{1}
\newline

@@

<<cache=TRUE>>=
X <- iris[1:4]; v <- as.character(iris$Species)
f <- qF(v, na.exclude = FALSE) # Adds 'na.included' class: no NA checks
gv <- group(v) # 'qG' object: first appearance order, with 'na.included'
microbenchmark(fmode(X, v), fmode(X, f), fmode(X, gv), fmode(X, g))
@
\vspace{-1mm}

\hrrule
\vspace{-2mm}
\section{Quick Conversions}
\itxt{Fast and exact conversion of common data objects} \\ [0.5em]
\code{qM(), qDF(), qDT(), qTBL()} - convert vectors, arrays, data.frames or lists to matrix, data.frame, data.table or tibble\\ [0.5em]
\code{m[r|c]tl()} - matrix rows/cols to list, data.frame or data.table\\ [0.5em]
\code{qF(), as\_numeric\_factor(), as\_character\_factor()} - convert to/from factors or all factors in a list / data.frame

@@

\section{Fast Data Manipulation}
\vspace{-1.4mm}
\itxt{Minimal overhead implementations} \newline
\setstretch{1.5}
\code{fselect[<-]()} - select/replace columns\\
\code{fsubset()} - subset data (rows and columns)\\
\code{ss()} - fast alternative to \code{[}, particularly for data frames\\
\code{[row|col]order[v]()} - reorder (sort) rows and columns\\
\code{fmutate(), fsummarise()} - \emph{dplyr}-like, incl. \code{across()} feature\\
\code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
\code{fcompute[v]()} - compute new cols dropping existing ones\\
\code{[f|set]rename()} - rename (any object with 'names' attribute)\\
\code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
\code{get\_vars[<-]()} - select/replace columns (standard eval.)\\ [0.5em]
\setstretch{1}
\code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
\code{add\_vars[<-]()} - add or column-bind columns\\
\code{rowbind()} - row-bind lists / data frame-like objects\\ [0.5em]
\code{join(), pivot()} - join and reshape data frame-like objects \newline

\textbf{Examples}
<<>>=
mtcars |> fsubset(mpg > fnth(mpg, 0.95), disp:wt, cylinders = cyl)
mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
i <- base::invisible # These are equivalent, the second option is faster:
mtcars |> fgroup_by(cyl, vs, am) |> fmutate(sum_mpg = fsum(mpg)) |> i()
mtcars |> fmutate(sum_mpg = fsum(mpg, list(cyl, vs, am), TRA = 1)) |> i()
# These are also equivalent (weighted means), again the second is faster
mtcars |> fgroup_by(cyl) |> fmutate(across(disp:drat, fmean, wt)) |> i()
mtcars |> ftransformv(disp:drat, fmean, cyl, wt, 1, apply = FALSE) |> i()
# ftransform()/fcompute() support list input and ignore attached groupings
mtcars %>% fgroup_by(cyl) %>% ftransform(fselect(., hp:qsec) %>%
           fmedian(TRA = 1) %>% fungroup() %>% fsum(TRA = "/")) |> i()
# Again a faster equivalent: note the use of 'set' to avoid a deep copy
mtcars %>% ftransform(fselect(., hp:qsec) %>% fmedian(cyl, TRA = 1) %>%
                      fsum(TRA = "/", set = TRUE)) %>% i()
# Aggregation: weighted standard deviations
mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
# Grouped linear models (one way of doing it)
qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(reg = list(lm(disp ~ drat)))
# Adding some columns. Use ftransform<- to also replace existing ones
add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("perc_")
@
<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
\vspace{-2mm}

@@

<<>>=
# Population weighted mean (PCGDP, LIFEEX) & mode (country), and sum(POP)
collap(wlddev, country + PCGDP + LIFEEX ~ income, w = ~ POP)
@
%\vspace{-2mm}
\end{multicols} % \vspace{-20mm}
%\vspace{20mm}

% \end{adjustbox}
%}
% \hrrule
\vspace{-5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.0\ \textbullet\ Updated: 2023-10
}

\newpage

% ------------------------------------------------------------------
% Second Page
% ------------------------------------------------------------------
\begin{multicols}{4}

@@

\itxt{Namespace masking}\\ [0.5em]
Can set \code{option(collpse\_mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R or \emph{dplyr}. A few keywords exist to mask multiple functions, see \code{help("collapse-options")}. This allows clean \& fast code, but poses additional namespace challenges:

@@

\end{multicols}

\vspace{-5.5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 1.8.8\ \textbullet\ Updated: 2022-08
}

\end{document}
